fGFPCA dynamic prediction (In-sample)

### 2023-01-04 f-GFPCA.Rmd
try MLE of out-of-sample latent function

@@

Now let's say if we have a new subject $\boldsymbol{Y}_{new}$ with observations up to $t_m$ and $t_m$ is in mth bin. Also, we do not wish to refit any model, including the local GLMMs. In the first step, we need to use the series of local GLMMs that's already been fitted to get the random intercepts for this new observations ($\boldsymbol{b}_{new}$).

We first focus on one specific interval s.

@@

```{r}
```

>>>>

## Some clarification points

1. When fitting local GLMMs, it seems that the intervals are essentially independent. We do not consider the effect of neighborhood intervals on a specific interval.

@@

Now let's say if we have a new subject $\boldsymbol{Y}$ with observations up to $t_m$ and $t_m$ is in mth bin. I can think of a few ways to  predict the following track, as below:

### Refit local GLMMs

@@

### Do not refit GLMMs

We can utilize the fitted series of local GLMMs, instead of refitting them. Since $\boldsymbol{Y}_{new}$ is a binary function, we can get its (incomplete) latent function value $\eta^s$ with Maximum Likelihood Estimation. Specifically, we need to estimate its (incomplete) random intercept function $b^s$.

Let's focus on a specific bin $s$. If the new sample has $n_s$ observations in this bin, among which $m_s$ are success (value = 1). Then $\boldsymbol{Y}^s$ would follow a Binomial distribution:

$$\boldsymbol{Y}^s \sim Binomial(n_s, \frac{exp(\eta^s)}{1+exp(\eta^s)})$$
Where $\eta^s = \beta_0^s+b^s$, and $\beta_0^s$ is already estimated with training data.

Now we write out the likelihood in interval s:

$$L(b^s|\boldsymbol{Y}^s) = \binom{n^s}{m^s}p_s^{m^s}(1-p_s)^{n^s-m^s}$$
where

$$p_s = \frac{exp(\eta^s)}{1+exp(\eta^s)}, \hspace{0.5cm} 
\eta^s = \beta_0^s+b^s$$

Then the log likelihood is:

$$\begin{aligned}
l(b^s|\boldsymbol{Y}^s) &= log\binom{n^s}{m^s}+{m^s}logp_s+(n^s-m^s)log(1-p_s)\\
&= log\binom{n^s}{m^s}+m^slog(\frac{p_s}{1-p_s})+n^slog(1-p_s)\\
&= log\binom{n^s}{m^s}+m^s(\beta_0+b^s)-n^slog(1+exp(\beta_0^s+b^s))
\end{aligned}$$

Then we take the first derivative and set it to zero, and end up with MLE os $b^s$:

$$\hat{b}^s=log(\frac{m^s}{n^s-m^s})-\beta_0^s$$
Again, $\beta_0^s$ is the estimated mean latent function from the local GlMMs steps. In fact, this individual latent track can be simply estimated by success ratio in the corresponding interval:

$$\eta^s = log\frac{m^s}{n^s-m^s}$$

```{r}
# generate one subject from the smae generating mechanism
xi_new <- matrix(rnorm(1*K),1,K);
xi_new <- xi_new %*% diag(sqrt(lambda))
# random intercept function
b_new <- xi_new %*% t(phi)
# latent function
eta_new <- f_0(sind) + b_new
## response binary function
Y_new <- rbinom(length(eta_new), size=1, prob=plogis(eta_new))
df_new_true <- data.frame(
  id = N+1000,
  sind = sind, 
  Y = Y_new,
  eta = t(eta_new),
  sind_inx = 1:J)

# cut observations after specific time point
tm <- 0.5
df_new_obs <- df_new_true %>%  select(-eta) %>% 
  filter(sind<=0.5)
```

```{r}
# bin observations
df_new_obs$bin <- cut(df_new_obs$sind_inx, breaks = brks, 
                      include.lowest = T,labels = mid)
df_new_obs$bin <- as.numeric(as.character(df_new_obs$bin))

# estimated MLE of latent function
eta_hat_new <- df_new_obs %>% group_by(bin) %>% 
  summarise_at("Y", function(x){log(sum(x==1)/sum(x==0))})

# plot
df_new_true %>% filter(sind_inx %in% mid) %>% 
  select(sind_inx, eta) %>% 
  rename(bin = sind_inx) %>% 
  left_join(eta_hat_new) %>% 
  ggplot()+
  geom_line(aes(x=bin, y=eta, col = "True"))+
  geom_line(aes(x=bin, y=Y, col = "Predicted"))
